Experiences Gamifying Developer Adoption\\ of Practices and Tools

### 2014-03-06 paper_id51.tex
Changed bib file location
some updates based on review

Like Sascha, software developers who maintain code face challenges of navigating an existing code base to incorporate changes as they build new features or fix bugs.  Toolsmiths respond to these challenges by creating new tools for assisting developers through difficult activities like code navigation.  However, as new practices and tools become available, developers may be motivated to start using them unless they interact with a peer developer who promotes them\cite{wbsnipes:Hill2011Peer}.    Of course they are intrinsically motivated to improve, but urgent demands interfere with experimenting using new tools and they become less efficient than otherwise possible.  Thus, we need to both communicate with and motivate developers like Sascha to start using new tools.

@@

\subsection{Monitoring Practice Studies }

@@

Johnson and Kou defined Zorro\cite{V:Johnson2007Automated}, a system for detecting whether developers use Test Driven Development techniques based on data from Hackystat.  Zorro divides development activities into episodes delimited by events such as configuration management code check-in, start of a unit test run, or start of a build.  Using the distinct events developers follow within these episodes, Zorro determines whether the episode followed Test-Driven Development practices per their classic definition of test first - code - test pass, or a different scenario.  In two student-based studies comparing Zorro classifications with a simultaneous observational screen video, Zorro achieved between 70\% \cite{Kou2010Operational} and 89\% \cite{V:Johnson2007Automated} accuracy when classifying episodes into their proper TDD scenarios.  The studies did not, however, attempt to use or evaluate the influence of instant feedback aimed at encouraging participants to follow the classic definition of Test Driven Development.  In our study, we attempt to present data collected directly to developers in order to influence them to change their practices.

@@

The Pro Metrics (PROM) tool provides a framework for collecting data for further analysis from tools used by developers.\cite{Coman2009Casestudy}  It provides a flexible data model and a plug-in architecture to facilitate collection from different data sources.  Studies conducted using PROM include a series of studies on trends in time spent Pair Programming \cite{Coman2008Investigating}, benefits of refactoring on productivity \cite{Moser2008Case}, impact of refactoring on re-usability \cite{Moser2006Does}, and prediction of effort \cite{Abrahamsson2007Effort}.  Studies using the PROM tool correlate the time spent editing with code metrics obtained through source code analysis.  Applying these data to specific research questions ,such as whether refactoring improves productivity \cite{Moser2008Case}, shows the utility of combining automated effort data with code metric data.  Taking a different focus on techniques, our study used more detailed events captured from the IDE  to detect commands the uses to navigating through the code in addition to knowing duration of edit sessions for code modules.

Murphy-Hill et al. study a large usage history data set and apply several different algorithms to accurately suggest IDE commands to novice users\cite{MurphyHill2012Improving}.  Algorithms based on command history performed well particularly when synchronized chronologically with the recipient's usage history. Novice users were better served by recommendation algorithms that did not require a long history, while more expert users benefited from more sophisticated algorithms that included more history. The ``Most Widely Used'' algorithm recommends commands based on the collective usage profile of the team and performed nearly as well as the more sophisticated algorithms. The paper presents the methodology for making recommendations and evaluating their accuracy. Our work takes a further step evaluating how well tools can influence developers to use a specific recommended commands.

@@

When discussing gamification, folks raise a concern about replacing an activity like software development, which is intrinsically motivated, with extrinsic motivation provided by points and achievements.  To discover what is behind this concern, we reviewed several studies on motivation and gamification.

@@

Maehr proposes an affirmative theory whereby  individuals achieve as a member of a social group, choosing the behavior that meets the expectations and values of the group that is significant to them\cite{wbsnipes:MaehrCulture}.  A key factor in raising achievement motivation is establishing a social group where the person is motivated to belong and excel.  Maehr states that since achievement behaviors can be triggered by circumstance, they are present in most people regardless of background.

Hamari et al. analyze achievements patterns in game systems and provide guidance on defining successful achievement systems.  They recommend  achievement awards be setup as a secondary form of scoring so players have multiple objectives available for a particular game.  Players should feel like they earned achievements in the game; they should not be given for trivial tasks.

Singer and Schneider demonstrate the use of a message board and points for encouraging students to increase their frequency of commits to the source code repository \cite{Singer2012It}.  The communication mechanism enabled students to see each others' progress and resulted in more frequent commits than baseline.  Participants valued the communication and collaboration aspect and some valued the competition enough to change their opinion on optimum commit frequency.  The subsequent thesis by Singer \cite{Singer2013a} describes results of an experiment conducted across two iterations of a class where active feedback on commits was deployed to one course and another course served as the control group where commit frequency was simply monitored.  Results show an increase in the frequency of commits at a statistically significant level.  This study uses fine grained instrumentation to identify practices used between commits such as structured navigation.

Dubois and Tamburrelli discuss theoretical concepts behind applying gamification to software development \cite{Dubois2013Understanding}. Their approach is to consider how gamification could be applied throughout the software development life-cycle to motivate many practices in different phases.  They provide guidance on assessing gamification in the software development domain including analyzing the actors and game mechanics, defining integration with existing tools and processes, and designing the evaluation of results.  Dubois and Tamburrelli demonstrate this guidance using a study of students working on a class project.  Defining a control group and subject group of students, both groups receive the same feedback about their code comments, test coverage, code size, and static analysis rule violations.  The subject group can see each others' rankings while the control group only sees their personal rankings.  Results, though inconclusive, show competition increases the desired results.  In our study, we provide participants with both individual feedback and anonymized competitive feedback and seek to observe a change in their behavior.

@@

Comments from participants reflected a more interesting controversial reaction than these measurements might otherwise indicate.  A developer from India said ``Gamification is a juvenile way of getting programmers to do something that you want.''  A different respondent expressed a concern that ``Gamification of best practices can have nasty side effects.''  Another participant said ``there is already high a level of competition between software development teams..'' indicating they are under considerable pressured to get the work done.  Another developer reacted with a statement that identifies the kind of intrinsic motivation we hope for ``I am using practices and tools because I want to do my job well',' indicating that gamification was unnecessary to motivate people to improve.

@@

The team goal ranked second behind coworker recommendation.  This response indicated that developers would be motivated to use new tools and practices if that helped their team in a competion between development teams at ABB.  The result shows that team competition would be first or second on the list for  47\% of developers to motivate their use of practices or tools.

@@

Following our data analysis, we conducted one on one feedback sessions with the pilot participants. Before initiating the discussion, we assured the individual developer that the data shared with them is confidential and we will not share the same with management. The participants received the feedback in the form of one-on one presentations of quantitative charts and written analysis of the information.  The key quantitative chart shown in Figure \ref{fig:developercomparison} compared their distribution of time in each category to the median for the group.

We discovered during these feedback sessions that developers did not spend the entire time in their desktop environment because they had a virtual machine environment available for testing. Thus categories like testing and debugging were affected by whether they used the virtual environment or their desktop for these activities.  Because those categories are not part of navigation, they are not discussed here. The developers use the virtual environment  explains some instances where developers have no activity recorded on some days.

@@

\section{Threats to Validity}
The threats to construct validity occur when measurements may not reflect the intended operational definition.The key measures of navigation ratio and points in our study could have results differing from their definition.

Measuring navigation ratio involved categorizing events into structured and unstructured navigation.  The categorization of events is specific to this study and documented herein, however, it may not be accurate to the intent of the operational definition of structured navigation depending on how the developer uses some commands like Find.  Another factor affecting this measurement may occur when developers repeat a command such as hitting the down arrow key many times, that may be something other than navigation (though it is classified as such for this study).  In feedback, we learned developers were reviewing and reading code which may account for these repetitive actions.

@@

%
% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
\balance
\bibliography{../mousetap} 
 % sigproc.bib is the name of the Bibliography in this case
% You must have a proper ".bib" file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references
%
% ACM needs 'a single self-contained file'!
%

>>>>

Like Sascha, software developers who maintain code face challenges of navigating an existing code base to incorporate changes as they build new features or fix bugs.  Toolsmiths respond to these challenges by creating new tools for assisting developers through difficult activities like code navigation.  However, as new practices and tools become available, developers may be motivated to start using them unless they interact with a peer developer who promotes them \cite{wbsnipes:Hill2011Peer}.    Of course they are intrinsically motivated to improve, but urgent demands interfere with experimenting using new tools and they become less efficient than otherwise possible.  Thus, we need to both communicate with and motivate developers like Sascha to start using new tools.

@@

\subsection{Monitoring Practice Studies}

@@

Johnson and Kou defined Zorro \cite{V:Johnson2007Automated}, a system for detecting whether developers use Test Driven Development techniques based on data from Hackystat.  Zorro divides development activities into episodes delimited by events such as configuration management code check-in, start of a unit test run, or start of a build.  Using the distinct events developers follow within these episodes, Zorro determines whether the episode followed Test-Driven Development practices per their classic definition of test first - code - test pass, or a different scenario. In two student-based studies comparing Zorro classifications with a simultaneous observational screen video, Zorro achieved between 70\% \cite{Kou2010Operational} and 89\% \cite{V:Johnson2007Automated} accuracy when classifying episodes into their proper TDD scenarios.  The studies did not, however, attempt to use or evaluate the influence of instant feedback aimed at encouraging participants to follow the classic definition of Test Driven Development.  In our study, we attempt to present data collected directly to developers in order to influence them to change their practices.

@@

The Pro Metrics (PROM) tool provides a framework for collecting data for further analysis from tools used by developers.\cite{Coman2009Casestudy}  It provides a flexible data model and a plug-in architecture to facilitate collection from different data sources.  Studies conducted using PROM include a series of studies on trends in time spent Pair Programming \cite{Coman2008Investigating}, benefits of refactoring on productivity \cite{Moser2008Case}, impact of refactoring on re-usability \cite{Moser2006Does}, and prediction of effort \cite{Abrahamsson2007Effort}.  Studies using the PROM tool correlate the time spent editing with code metrics obtained through source code analysis.  Applying these data to specific research questions, such as whether refactoring improves productivity \cite{Moser2008Case}, shows the utility of combining automated effort data with code metric data.  Taking a different focus on techniques, our study used more detailed events captured from the IDE to detect commands used to navigating through the code in addition to knowing duration of edit sessions for code modules.

Murphy-Hill et al. study a large usage history data set and apply several different algorithms to accurately suggest IDE commands to novice users\cite{MurphyHill2012Improving}.  Algorithms based on command history performed well particularly when synchronized chronologically with the recipient's usage history. Novice users were better served by recommendation algorithms that did not require a long history, while more expert users benefited from more sophisticated algorithms that included more history. The ``Most Widely Used'' algorithm recommends commands based on the collective usage profile of the team and performed nearly as well as the more sophisticated algorithms. The paper presents the methodology for making recommendations and evaluating their accuracy. Our work takes a further step evaluating how well tools can influence developers to use specific recommended commands.

@@

When discussing gamification, concerns are raised about replacing an activity like software development, which is intrinsically motivated, with extrinsic motivation provided by points and achievements.  To discover what is behind this concern, we reviewed several studies on motivation and gamification.

@@

Maehr proposes an affirmative theory whereby individuals achieve as a member of a social group, choosing the behavior that meets the expectations and values of the group that are significant to them\cite{wbsnipes:MaehrCulture}.  A key factor in raising achievement motivation is establishing a social group where the person is motivated to belong and excel.  Maehr states that since achievement behaviors can be triggered by circumstance, they are present in most people regardless of background.

Hamari et al. analyze achievements patterns in game systems and provide guidance on defining successful achievement systems.  They recommend achievement awards be setup as a secondary form of scoring so players have multiple objectives available for a particular game.  Players should feel like they earned achievements in the game instead given trivial tasks.

Singer and Schneider demonstrate the use of a message board and points for encouraging students to increase their frequency of commits to the source code repository \cite{Singer2012It}.  The communication mechanism enabled students to see each other's progress and resulted in more frequent commits than baseline.  Participants valued the communication and collaboration aspect and some valued the competition enough to change their opinion on optimum commit frequency.  The subsequent thesis by Singer \cite{Singer2013a} describes results of an experiment conducted across two iterations of a class where active feedback on commits was deployed to one course and another course served as the control group where commit frequency was simply monitored.  Results show an increase in the frequency of commits at a statistically significant level.  This study uses fine grained instrumentation to identify practices used between commits such as structured navigation.

Dubois and Tamburrelli discuss theoretical concepts behind applying gamification to software development \cite{Dubois2013Understanding}. Their approach is to consider how gamification could be applied throughout the software development life-cycle to motivate many practices in different phases.  They provide guidance on assessing gamification in the software development domain including analyzing the actors and game mechanics, defining integration with existing tools and processes, and designing the evaluation of results.  Dubois and Tamburrelli demonstrate this guidance using a study of students working on a class project.  Defining a control group and subject group of students, both groups receive the same feedback about their code comments, test coverage, code size, and static analysis rule violations.  The subject group can see each other's rankings while the control group only sees their personal rankings.  Results, though inconclusive, show competition increases the desired results.  In our study, we provide participants with both individual feedback and anonymized competitive feedback and seek to observe a change in their behavior.

@@

Comments from participants reflected a more interesting controversial reaction than these measurements might otherwise indicate.  A developer from India said ``Gamification is a juvenile way of getting programmers to do something that you want.''  A different respondent expressed a concern that ``Gamification of best practices can have nasty side effects.''  Another participant said ``there is already high a level of competition between software development teams.'' indicating they are under considerable pressured to get the work done.  Another developer reacted with a statement that identifies the kind of intrinsic motivation we hope for ``I am using practices and tools because I want to do my job well',' indicating that gamification was unnecessary to motivate people to improve.

@@

The team goal ranked second behind coworker recommendation.  This response indicated that developers would be motivated to use new tools and practices if that helped their team in a competition between development teams at ABB.  The result shows that team competition would be first or second on the list for 47\% of developers to motivate their use of practices or tools.

@@

Following our data analysis, we conducted one-on-one feedback sessions with the pilot participants. Before initiating the discussion, we assured the individual developer that the data shared with them is confidential and we will not share the same with management. The participants received the feedback in the form of one-on one presentations of quantitative charts and written analysis of the information.  The key quantitative chart shown in Figure \ref{fig:developercomparison} compared their distribution of time in each category to the median for the group.

We discovered during these feedback sessions that developers did not spend the entire time in their desktop environment because they had a virtual machine environment available for testing. Thus categories like testing and debugging were affected by whether they used the virtual environment or their desktop for these activities.  Because those categories are not part of navigation, they are not discussed here. Due to this, there are some instances where developers have no activity recorded on some days.

@@

\section{Threats to Validity}
The threats to construct validity occur when measurements may not reflect the intended operational definition. The key measures of navigation ratio and points in our study could have results differing from their definition.

Measuring navigation ratio involved categorizing events into structured and unstructured navigation.  The categorization of events is specific to this study and documented herein, however, it may not be accurate to the intent of the operational definition of structured navigation depending on how the developer uses some commands like Find.  Another factor affecting this measurement may occur when developers repeat a command such as hitting the down arrow key many times, which may be something other than navigation (though it is classified as such for this study).  In feedback, we learned developers were reviewing and reading code which may account for these repetitive actions.

@@

%
% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
\balance
\bibliography{mousetap} 
 % sigproc.bib is the name of the Bibliography in this case
% You must have a proper ".bib" file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references
%
% ACM needs 'a single self-contained file'!
%
